---
title: "Assignment0 : Reproducible Research and Exploratory Data Analysis in R"
author: "Mohshinul Karim"
date: "2024-05-17"
output: 
  html_document: 
    fig_width: 8
    fig_height: 8
    toc: yes
    theme: journal
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
x<-2
```

```{r load-packages, warning=FALSE}
#Load these packages

library(tidyverse) 
library(datasauRus)
warning=FALSE
```

#Data 

**1. Based on the help file, how many rows and how many columns does the datasaurus_dozen file have? What are the variables included in the data frame? (this can be hardcoded)**

__Answer: A data frame with 1846 rows and 3 Columns. Variables: dataset, x, y.__


```{r}
# Make a frequency table of the 'dataset' variable to look at what these datasets:
datasaurus_dozen %>%
  count(dataset)
```
#Data visualization and summary

**2. Plot y vs. x for the dino dataset. Then, calculate the correlation coefficient between x and y for this dataset**

```{r}
#Start with the datasaurus_dozen and pipe it into the filter function to filter for observations where dataset == "dino". Store the resulting filtered data frame as a new data frame called dino_data.
dino_data <- datasaurus_dozen %>%
  filter(dataset == "dino")
```


```{r fig.height=5, fig.width=9}
#visualizing these data
ggplot(data = dino_data, mapping = aes(x = x, y = y)) +
  geom_point()
```

**Start with dino_data and calculate a summary statistic that we will call r as the correlation between x and y**
```{r}
dino_data %>%
  summarize(r = cor(x, y))
```
**3. Plot y vs. x for the star dataset. You can (and should) reuse code we introduced above, just replace the dataset name with the desired dataset. Then, calculate the correlation coefficient between x and y for this dataset. How does this value compare to the r of dino?**


```{r}
#Create Star_data data frame
star_data <- datasaurus_dozen %>%
  filter(dataset == "star")
```

**Plot:**
```{r fig.height=5, fig.width=5}
ggplot(data = star_data, mapping = aes(x = x, y = y)) +
  geom_point()
```

**Summarize:**
```{r}
star_data %>%
  summarize(r = cor(x, y))
```
**Comparison: Dino Coefficient: -0.06447185	and star coefficient: -0.0629611. the absolute value is lower for star_data.** 

**4. Plot y vs. x for the circle dataset. You can (and should) reuse code we introduced above, just replace the dataset name with the desired dataset. Then, calculate the correlation coefficient between x and y for this dataset. How does this value compare to the r of dino?**


```{r}
#Create circle_data data frame
circle_data <- datasaurus_dozen %>%
  filter(dataset == "circle")
```

**Plot:** 
```{r fig.height=5, fig.width=5}
ggplot(data = circle_data, mapping = aes(x = x, y = y)) +
  geom_point()
```

**Summarize:**
```{r}
circle_data %>%
  summarize(r = cor(x, y))
```

**comparison: Dino Coefficient: -0.06447185	and circle coefficient: -0.06834336. the absolute value is higher for circle_data.** 



**5. Finally, let’s plot all datasets at once. In order to do this we will make use of facetting**
```{r fig.height=12, fig.width=10}
# Create the faceted plot
ggplot(datasaurus_dozen, aes(x = x, y = y, color = dataset))+
  geom_point()+
  facet_wrap(~ dataset, ncol = 3) +
  theme(legend.position = "none") +
   labs(title = "Datasaurus Dozen Datasets", x = "X", y = "Y")
```


```{r}
#using the group_by function to generate all correlation coefficients. 
# Defining r as the correlation coefficient

datasaurus_dozen %>%
  group_by(dataset) %>%
  summarize(r = cor(x, y))
```


It was Fun!!!
                          
<mark> Thank YOU </mark>.





